Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs

### 2025-03-30 ResponseLetter2.tex

\textcolor{red}{Recurrence time is a increasing function of sensitive and aneuploid growth rates not a decreasing function as stated in the manuscript. Change RS2}

\textcolor{red}{"anuploids"$\rightarrow$"aneuploids". Change RS3}

\textcolor{red}{"aneuplidies"$\rightarrow$"aneuploids". Change RS4}

\textcolor{red}{$1/u\lambda_a$ is not the expected time for an aneuploid lineage to appear and we have corrected the interpretation of $\abs{\Delta_a}/u\lambda_a$ as it is the ratio of the number of sensitive births required to produce an aneuploid, $1/u$, to the number of births in the resulting aneuploid lineage before it goes extinct, $\lambda_a / |\Delta_a|$. Change RS5}

\textcolor{red}{We deleted the sentence "Drugs that significantly increase the division rate of sensitive cells, $\lambda_s$, but do not affect their death rate, $\mu_s$, decrease cancer recurrence times (conditioned on evolutionary rescue)" because while it is mathematically true it not make sense for the drug to increase the fitness of sensitive cells. Change RS6}

### 2025-04-20 ms.tex
revised responses

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Evolutionary rescue occurs when a population, facing a sudden environmental change that would otherwise lead to extinction, adapts through beneficial mutations, allowing it to recover and persist.  % change 32
A prime example of evolutionary rescue is the ability of cancer to survive exposure to treatment. 
One evolutionary mechanism by which a population of cancer cells can adapt to chemotherapy is aneuploidy. 
Aneuploid cancer cells can be more fit in an environment altered by anti-cancer drugs, because aneuploidy disrupts the pathways usually targeted by the drugs. %change 3 %change S2
Indeed, aneuploidy is highly prevalent in tumors, and some anti-cancer drugs fight cancer by increasing chromosomal instability.
Here, we model the impact of aneuploidy on the fate of a population of cancer cells. We use multi-type branching processes to approximate the probability that a tumor survives drug treatment as a function of the initial tumor size, the rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the drug-sensitive and drug-resistant cells. We also investigate the effect of the pre-existent aneuploid cells on the probability of evolutionary rescue. %change S2
 Finally, we estimate the tumor's mean recurrence time to revert to its initial size following treatment and evolutionary rescue.
We propose that aneuploidy can play an essential role in the relapse of smaller secondary tumors.
\end{abstract}

@@

\paragraph{Evolutionary rescue.} Populations adapted to a specific environment are vulnerable to environmental changes, which might cause the population's extinction. Examples of such environmental changes include climate change, invasive species, and the onset of drug therapies. Adaptation is a race against time as the population size decreases in the new environment~\citep{tanaka2022surviving}. 
\emph{Evolutionary rescue} is the process by which the population acquires an adaptation that increases fitness in the new environment such that extinction is averted.
There are three potential ways for a population to survive environmental change \citep{alexander2014evolutionary,bell2017evolutionary}: % change 45 %change 28
migration to a new habitat similar to the one before the onset of environmental change \citep{harsch2014keeping,cobbold2020should,zhou2022range}; 
adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non,gunnarsson2020understanding}; 
and adaptation through genetic modifications, like mutation \citep{gomulkiewicz1995does,uecker2014evolutionary,uecker2016role,uecker2011fixation,orr2014population}. %change S5
Here, we focus on the latter.

@@

We will address these questions essentially by combining known results from the evolutionary rescue literature, and considering the quantities and parameter values that are relevant for cancer.
Because these results are scattered across multiple papers, and because they all follow from simple branching process approximations, we re-derive all results here for our model.
While our focus is on the application of evolutionary rescue theory to the evolution of drug resistance in cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available, as in \citet{iwasa2003evolutionary, iwasa2004evolutionary} and \citet{osmond_genetic_2020}.

@@

\paragraph{Evolutionary model.}
We follow the number of cancer cells with one of three different genotypes at time $t$: sensitive, $s_t$; tolerant/resistant aneuploid, $a_t$; and resistant mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=s, a, m$).
The division and death rate difference is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, and therefore $\Delta_s<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
We consider a range of possible values for $\Delta_a$, finding three distinct scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_s$ \citep[tolerant cells are able to survive the effects of the drug for an extended period of time compared to susceptible cells while resistant cells require a higher drug concentration to be affected, see][for a detailed explanation of the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary, or growing or dying slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below. % change S1 %change S17

@@

Note that when we refer to drug-sensitive cells, we include those cells that have any aneuploidy that does not affect fitness in the presence of the drug (see \Cref{table1}). %change S7
Additionally, we focus on mutations that confer resistance, neglecting deleterious mutations (which are more common). We assume deleterious mutations and other aneuploidies occur at similar rates across genotypes and therefore neglect their effect on the dynamics. %change RS4

@@

Using \cref{eq:N_a,eq:N_m}, we can find the ratio of threshold tumor size for rescue via aneuploidy ($u$ is high) or via direct mutation ($u$ is low),
\begin{equation} \label{eq:N_ratio}
\frac{N^*_a}{N^*_m} \approx \begin{cases}
    \frac{\abs{\Delta_a}}{u\lambda_a} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)} ,\\ 
  \frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1 \text{ (stationary aneuploids)} ,\\ 
  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)} .
  \end{cases}
\end{equation}
Imporantly, when this threshold is smaller than 1, aneuploidy is expected to play a role in evolutionary rescue, as it reduces the threshold tumor size neeede for rescue. %change S8
As expected, this ratio increases with the mutation rate $v$ and decreases with the aneuploidy rate $u$.
In the first scenario, $\abs{\Delta_a}/u\lambda_a$ is the ratio of the number of sensitive divisions required to produce an aneuploid, $1/u$, to the number of divisions in the resulting aneuploid lineage before it goes extinct, $\lambda_a / |\Delta_a|$.
% RS5
%is the ratio of the expected time for an aneuploid lineage to appear, $1/u\lambda_a$, and the expected time until that lineage disappears, $1/\abs{\Delta_a}$.
In the third scenario, $\left(v \frac{\Delta_m}{\lambda_m}\right) / \left(u \frac{\Delta_a}{\lambda_a}\right)$ is the ratio of the rates of appearance of resistant mutants that avoid extinction and partially resistant aneuploids that avoid extinction.
In the second scenario, $\frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2}=\sqrt{\frac{\Delta_a}{u\lambda_a}  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}}$, which is the geometric mean of the first and third scenarios.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph*{Standing vs. \textit{de novo} genetic variation.}

We assumed that at the onset of drug treatment the initial tumor consisted entirely of drug-sensitive cells but aneuploidy is likely produced even before the onset of treatment at some rate $\tilde{u}$, which may be lower in the absence of drugs, $\tilde{u} < u$ \citep{wang2019molecular,mason2017functional}. %change S9, S19 
 Aneuploidy likely confers a fitness cost $c$ in the absence of drugs \citep{replogle2020aneuploidy,giam2015aneuploidy}. %change S16, S19 
If the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_s/c$ of aneuploid cells in the population (here we assume that the drug affects the sensitive death rate but not the division rate and therefore we use $\lambda_s$ for the sensitive division rate in the drug-free environment). %change S16, S19

@@

Using \cref{eq:N_a,eq:N_m,eq:Nsgv}, we can find the ratio of threshold tumor size for rescue via standing genetic variation  to the threshold for rescue via direct mutation,
\begin{equation} \label{eq:N_sgv_ratio}
\frac{\tilde{N}^*_a}{N^*_m}=\frac{\tilde{N}^*_{a}}{N^*_{a}} \frac{N^*_a}{N^*_m} \approx \frac{c}{\abs{\Delta_s}}\begin{cases}
    \frac{\abs{\Delta_a}}{\tilde{u}\lambda_a} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)} ,\\ 
  \frac{1}{\tilde{u}}\left(v \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1 \text{ (stationary aneuploids)} ,\\ 
  v \frac{\Delta_m}{\lambda_m} \left(\tilde{u}\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1\text{ (resistant aneuploids)} .
  \end{cases}
\end{equation}
Evolutionary rescue through direct mutation is more likely if the cost of aneuploidy, $c$, is large or the effect of the drug $\Delta_s$ is small. In contrast, standing genetic variation will drive adaptation if the pre-treatment chromosome missegregation rate, $\tilde{u}$, is large. % change S1
The ratio does not depend on the rate of chromosome missegregation induced by the drug, $u$. However, if the aneuploid growth rate, $\Delta_a$, increases, evolutionary rescue is driven by standing genetic variation. For the parameter values of  $\lambda_s=0.1$, $\lambda_a=0.0899$, $\lambda_m=0.1$, $\mu_s=0.14$, $\mu_a=0.09$, $\mu_m=0.09$, $\tilde{u}=10^{-3}$, and $v=10^{-7}$, we are in the first scenario (tolerant aneuploids) and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 1.94$, which means that standing genetic variation does not drive evolution of drug resistance when compared to direct mutation. We note that for larger values of the pre-treatment chromosome missegregation rate, $\tilde{u}$, which are consistent with empirical studies (\Cref{table1}), standing genetic variation can drive adaptation when compared to direct mutation.

@@

When evolutionary rescue occurs, the time until the tumor recurs may still be long. We therefore explored the time until the tumor recurs, that is, the time until the tumor reaches its original size, $N$. % change S1
When the expected number of resistant lineages that avoid extinction is small, the expected recurrence time can be estimated by adding two terms: the \emph{mean evolutionary rescue time}, which is the waiting time for the appearance of a resistant lineage that avoids extinction (conditioned on such an event occurring in the first place), and the \emph{mean proliferation time}, which is the expected time for that lineage to grow to $N$ cells.
However, when the expected number of resistant lineages is large, the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time because multiple mutant lineages contribute towards the mutant population size reaching the initial tumor size.
In this case the dynamics of the number of mutant cells is deterministic and can therefore be modeled by a system of ordinary differential equations (ODE), which describe how the number of mutant cells changes over time by its time derivative (eq.~\ref{detODE}). %change 2
Of particular interest is the distribution of the evolutionary rescue time and recurrence time with tolerant aneuploid cells ($\Delta_a T^*\ll1$) given that this represents the two step evolutionary rescue scenario. We focus on the parameter values in \Cref{table1} with $\lambda_a=0.0899$ (tolerant aneuploids), $\mu_s=0.14$ (intermediate sensitive death rate), and $v=10^{-7}$ (high end of the mutation rate). %change S11

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\begin{table}
\begin{center}
  \begin{tabular}{| l |p{5cm}| c | c | p{3cm} |}
    \hline
     & Name & Value & Units & References \\ \hline
    $N$ & Initial tumor size & $10^7-10^9$ & cells  & \citet{del2009does} \\ \hline
    $\lambda_s$ & Sensitive division rate& 0.1 & 1/days  & \citet{bozic2013evolutionary,rew2000cell} \\ \hline
    $\mu_s$ & Sensitive death rate& $0.11-0.17$ & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $\lambda_a$  & Aneuploid division rate$^\ast$ & $0.06-0.1$ & 1/days  & - \\ \hline
    $\mu_a$ & Aneuploid death rate$^\ast$ & $0.09$ & 1/days  & - \\ \hline
    $\lambda_m$ & Mutant division rate& 0.1 & 1/days  & \citet{bozic2013evolutionary,rew2000cell} \\ \hline
    $\mu_m$ & Mutant death rate& 0.09 & 1/days  & \citet{bozic2013evolutionary,carlson2003tumor} \\ \hline
    $u$ & Missegregation rate& $10^{-2}$ & 1$\slash$cell division  & \citet{lee2016effects} \\ \hline
    $v$ & Mutation rate& $10^{-9}-10^{-7}$ &  1$\slash$cell division  & \citet{bozic2013evolutionary,loeb2001mutator} \\  \hline
    $\tilde{u}$ & Missegregation rate in the drug free environment$^\ast$& $5\times10^{-4}-2\times10^{-2}$ & 1$\slash$cell division  & \citet{shi2005chromosome,thompson2008examining} \\ \hline
    $c$ & Selection coefficient against aneuploidy in the drug free environment& $0.07$ &  1/days   & \citet{lukow2021chromosomal} \\ \hline
  \end{tabular}
\caption{\textbf{Model parameters: Melanoma.} 
Parameters from \citet{bozic2013evolutionary} consider patients with melanoma treated with the anti-cancer drug vemurfenib, in which resistance is conferred by trisomy in either Chr 2 or Chr 6.
We have modified the parameters from \citet{bozic2013evolutionary} such that sensitive and mutant division rates are $\lambda_{s} = \lambda_{m} = \log{(2)}/T\approx0.1$ instead of their value of $0.14$ where $T$ is the doubling time in the absence of cellular death obtained from \citet{rew2000cell}. For a discussion of the different interpretations of the tumor doubling times see \citet{avanzini2019cancer}. Parameters marked with $^\ast$ are not obtained from the literature.} %change 1
  \label{table1}
\end{center}
\end{table}

@@

\begin{figure}
\begin{subfigure}{0.48\textwidth}
A\\
\includegraphics[width=1\textwidth]{Figures/ProbvNPlot.pdf}
\end{subfigure}
\begin{subfigure}{0.52\textwidth}
B\\
\includegraphics[width=1\textwidth]{Figures/PDXModelProb.pdf}
\end{subfigure}
\caption{\textbf{Aneuploidy facilitates the evolutionary rescue of cancer under drug treatment.}
The probability of evolutionary rescue, $\presc$, as a function of the initial tumor size, $N$ (\cref{eq:rescue_prob}). 
Dashed vertical lines show the threshold tumor size, $N_a^*$, above which the probability is high (\cref{eq:N_a}). % change S1
\textbf{(A)} Blue dashed line: without aneuploidy ($u=0$). Black line: tolerant aneuploidy ($u=10^{-2}, \lambda_a=0.0899$). Red line: stationary aneuploidy ($u=10^{-2}, \lambda_a=0.08999$). Green line: partially resistant aneuploidy ($u=10^{-2}, \lambda_a=0.095$). Markers and error bars for averages of simulation results with $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has been rescued, and $n=100$ is the number of simulations). Parameters: $\lambda_s=0.1,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7}$.
\textbf{(B)} Comparison for melanoma and breast cancer with parameters from the literature (\Cref{table1,table2}). Black line: Melanoma A375. Blue line: Breast cancer TNBC-SA1035 clone H. Red line: Breast cancer TNBC-SA609 clone A. Green line: Breast cancer TNBC-SA535 clone H.
}
\label{rescue_prob}
\end{figure}

@@

To analyze evolutionary rescue in our model, we use the framework of \emph{multi-type branching processes} \citep{harris1963theory, weissman2009rate}. 
In the case of cancer, where all reproduction is binary fission, these branching processes are also \emph{birth-death processes} \citep{kendall1948}% change 31
This allows us to find explicit expressions for the \emph{survival probability}: the probability that a lineage descended from a single cell does not become extinct.

@@

\caption{\textbf{Aneuploidy increases the number of mutations which rescue the tumor.} Shown is the expected number of mutations, which will rescue the cancer cell population, produced through the evolutionary trajectory \textit{sensitive} $\rightarrow$ \textit{mutant} (blue line, \cref{eq:twosteplineagedirect}) or through the trajectory \textit{sensitive} $\rightarrow$ \textit{aneuploid} $\rightarrow$ \textit{mutant} (red line, \cref{eq:twosteplineage}). Dashed vertical red line represents the threshold tumor size above which evolutionary rescue is likely through aneuploidy \cref{eq:N_a} and the dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is likely to occur through direct mutation \cref{eq:N_m}. % change S1

@@

\caption{The mean time for appearance of a resistance mutation that leads to evolutionary rescue with aneuploidy ($u>0$). 

@@

\caption{The mean time for the mutant cell population to reach size $N$, where $N$ is the initial number of cancer cells.

@@

\caption{The mean time for the mutant cell population to reach size $M$, where $M$ is the tumor detection size.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Evolutionary rescue occurs when a population, facing a sudden environmental change that would otherwise lead to extinction, adapts through beneficial mutations, allowing it to recover and persist.  % change 32
A prime example of evolutionary rescue is the ability of cancer to survive exposure to treatment. 
One evolutionary mechanism by which a population of cancer cells can adapt to chemotherapy is aneuploidy. 
Aneuploid cancer cells can be more fit in an environment altered by anti-cancer drugs, in part because aneuploidy may disrupt the pathways targeted by the drugs.
Indeed, aneuploidy is highly prevalent in tumors, and some anti-cancer drugs fight cancer by increasing chromosomal instability.
Here, we model the impact of aneuploidy on the fate of a population of cancer cells. We use multi-type branching processes to approximate the probability that a tumor survives drug treatment as a function of the initial tumor size, the rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the drug-sensitive and drug-resistant cells. We also investigate the effect of the pre-existent aneuploid cells on the probability of evolutionary rescue. %change S2
 Finally, we estimate the tumor's mean recurrence time to revert to its initial size following treatment and evolutionary rescue.
We propose that aneuploidy can play an essential role in the relapse of smaller secondary tumors.
\end{abstract}

@@

\paragraph{Evolutionary rescue.} Populations adapted to a specific environment are vulnerable to environmental changes, which might cause the population's extinction. Examples of such environmental changes include climate change, invasive species, and the onset of drug therapies. Adaptation is a race against time as the population size decreases in the new environment~\citep{tanaka2022surviving}. 
\emph{Evolutionary rescue} is the process by which the population acquires an adaptation that increases fitness in the new environment such that extinction is averted.
There are three potential ways for a population to survive environmental change \citep{alexander2014evolutionary,bell2017evolutionary}: % change 45 %change 28
migration to a new habitat similar to the one before the onset of environmental change \citep{harsch2014keeping,cobbold2020should,zhou2022range}; 
adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non,gunnarsson2020understanding}; 
and adaptation through genetic modifications like mutation \citep{gomulkiewicz1995does,uecker2014evolutionary,uecker2016role,uecker2011fixation,orr2014population}. %change S5
Here, we focus on the latter.

@@

We will address these questions essentially by combining known results from the evolutionary rescue literature, and considering the quantities and parameter values that are relevant for cancer.
Because these results are scattered across multiple papers, and because they all follow from simple branching process approximations, we re-derive all results here for our model.
While our focus is on the application of evolutionary rescue theory to the evolution of drug resistance in cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available, as in \citet{osmond_genetic_2020} and  \citet{iwasa2003evolutionary, iwasa2004evolutionary}.

@@

\paragraph{Evolutionary model.}
We follow the number of cancer cells with one of three different genotypes at time $t$: sensitive, $s_t$; tolerant/resistant aneuploid, $a_t$; and resistant mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=s, a, m$).
The division and death rate difference is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, and therefore $\Delta_s<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
We consider a range of possible values for $\Delta_a$, finding three distinct scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_s$; in the third, aneuploid cells are non-growing, stationary, or growing or dying slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below. % change S1 %change S17
We define tolerant cells as those that survive drug exposure longer than susceptible cells, while resistant cells require higher drug concentrations to be affected. For details on these distinctions, see \citet{brauner2016distinguishing}.

@@

Note that when we refer to drug-sensitive cells, we include those cells that have any aneuploidy that does not affect fitness in the presence of the drug (\Cref{table1}). %change S7
Additionally, we focus on mutations that confer resistance, neglecting deleterious mutations (which are more common). We assume deleterious mutations and other aneuploidies occur at similar rates across genotypes and therefore neglect their effect on the dynamics. %change RS4

@@

Using \cref{eq:N_a,eq:N_m}, we can find the ratio of threshold tumor size for rescue via aneuploidy ($u$ is high) or via direct mutation ($u$ is low),
\begin{equation} \label{eq:N_ratio}
\frac{N^*_a}{N^*_m} \approx \begin{cases}
    \frac{\abs{\Delta_a}}{u\lambda_a} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)} ,\\ 
  \frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1 \text{ (stationary aneuploids)} ,\\ 
  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)} .
  \end{cases}
\end{equation}
Imporantly, when this threshold is smaller than 1, aneuploidy is expected to play a role in evolutionary rescue, as it reduces the threshold tumor size needed for rescue. %change S8
As expected, this ratio increases with the mutation rate $v$ and decreases with the aneuploidy rate $u$.
In the first scenario, $\abs{\Delta_a}/u\lambda_a$ is the ratio of the number of sensitive divisions required to produce an aneuploid, $1/u$, to the number of divisions in the resulting aneuploid lineage before it goes extinct, $\lambda_a / |\Delta_a|$.
% RS5
%is the ratio of the expected time for an aneuploid lineage to appear, $1/u\lambda_a$, and the expected time until that lineage disappears, $1/\abs{\Delta_a}$.
In the third scenario, $\left(v \frac{\Delta_m}{\lambda_m}\right) / \left(u \frac{\Delta_a}{\lambda_a}\right)$ is the ratio of the rates of appearance of resistant mutants that avoid extinction and partially resistant aneuploids that avoid extinction.
In the second scenario, $\frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2}=\sqrt{\frac{\Delta_a}{u\lambda_a}  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}}$, which is the geometric mean of the first and third scenarios.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph*{Standing vs. \textit{de novo} genetic variation.}
We initially assumed that at the onset of drug treatment, the tumor consisted entirely of drug-sensitive cells. In reality, aneuploidy is likely generated even before treatment begins, at some rate $\tilde{u}$, which may be lower in the absence of drugs, $\tilde{u} < u$ \citep{wang2019molecular,mason2017functional}. %change S9, S19 
Aneuploidy is also thought to carry a fitness cost $c$ when drugs are absent \citep{replogle2020aneuploidy,giam2015aneuploidy}. %change S16, S19 
Therefore, if the tumor size $N$ is large--as expected at the time of treatment--a fraction of cells, approximately $f \approx \tilde{u}\lambda_s/c$, may already be aneuploid.
Here we assume that the drug affects the sensitive death rate but not the division rate and therefore we use $\lambda_s$ for the sensitive division rate in the drug-free environment. %change S16, S19

@@

Using \cref{eq:N_a,eq:N_m,eq:Nsgv}, we can find the ratio of threshold tumor size for rescue via standing genetic variation  to the threshold for rescue via \textit{de novo} genetic variation,
\begin{equation} \label{eq:N_sgv_ratio}
\frac{\tilde{N}^*_a}{N^*_m}=\frac{\tilde{N}^*_{a}}{N^*_{a}} \frac{N^*_a}{N^*_m} \approx \frac{c}{\abs{\Delta_s}}\begin{cases}
    \frac{\abs{\Delta_a}}{\tilde{u}\lambda_a} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)} ,\\ 
  \frac{1}{\tilde{u}}\left(v \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1 \text{ (stationary aneuploids)} ,\\ 
  v \frac{\Delta_m}{\lambda_m} \left(\tilde{u}\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1\text{ (resistant aneuploids)} .
  \end{cases}
\end{equation}
Evolutionary rescue through direct mutation is more likely if the cost of aneuploidy, $c$, is large or the effect of the drug $\Delta_s$ is small. In contrast, standing genetic variation will drive adaptation if the pre-treatment chromosome missegregation rate, $\tilde{u}$, is large. % change S1
The ratio does not depend on the rate of chromosome missegregation induced by the drug, $u$. However, if the aneuploid growth rate, $\Delta_a$, increases, evolutionary rescue is driven by standing genetic variation. For the parameter values of  $\lambda_s=0.1$, $\lambda_a=0.0899$, $\lambda_m=0.1$, $\mu_s=0.14$, $\mu_a=0.09$, $\mu_m=0.09$, $\tilde{u}=10^{-3}$, and $v=10^{-7}$, we are in the first scenario (tolerant aneuploids) and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 1.94$, which means that standing genetic variation does not drive evolution of drug resistance when compared to direct mutation. We note that for larger values of the pre-treatment chromosome missegregation rate, $\tilde{u}$, which are consistent with empirical studies (\Cref{table1}), standing genetic variation can drive adaptation when compared to direct mutation.

@@

When evolutionary rescue occurs, the time until the tumor recurs may still be long. We therefore explored the time until the tumor recurs, that is, the time until the tumor reaches its original size, $N$. % change S1
When the expected number of resistant lineages that avoid extinction is small, the expected recurrence time can be estimated by adding two terms: the \emph{mean evolutionary rescue time}, which is the waiting time for the appearance of a resistant lineage that avoids extinction (conditioned on such an event occurring in the first place), and the \emph{mean proliferation time}, which is the expected time for that lineage to grow to $N$ cells.
However, when the expected number of resistant lineages is large, the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time because multiple mutant lineages contribute towards the mutant population size reaching the initial tumor size.
In this case the dynamics of the number of mutant cells is deterministic and can therefore be modeled by a system of ordinary differential equations (ODE), which describe how the number of mutant cells changes over time by its time derivative (eq.~\ref{detODE}). %change 2
Of particular interest is the distribution of the evolutionary rescue time and recurrence time with tolerant aneuploid cells ($\Delta_a T^*\ll1$) as it represents the two-step evolutionary rescue scenario. We focus on the parameter values in \Cref{table1} with $\lambda_a=0.0899$ (tolerant aneuploids), $\mu_s=0.14$ (intermediate sensitive death rate), and $v=10^{-7}$ (high end of the mutation rate). %change S11

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\begin{table}
\begin{center}
  \begin{tabular}{| l |p{5cm}| c | c | p{3cm} |}
    \hline
     & Name & Value & Units & References \\ \hline
    $N$ & Initial tumor size & $10^7-10^9$ & cells  & \citet{del2009does} \\ \hline
    $\lambda_s$ & Sensitive division rate& 0.1 & 1/days  & \citet{bozic2013evolutionary,rew2000cell} \\ \hline
    $\mu_s$ & Sensitive death rate& $0.11-0.17$ & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $\lambda_a$  & Aneuploid division rate$^\ast$ & $0.06-0.1$ & 1/days  & - \\ \hline
    $\mu_a$ & Aneuploid death rate$^\ast$ & $0.09$ & 1/days  & - \\ \hline
    $\lambda_m$ & Mutant division rate& 0.1 & 1/days  & \citet{bozic2013evolutionary,rew2000cell} \\ \hline
    $\mu_m$ & Mutant death rate& 0.09 & 1/days  & \citet{bozic2013evolutionary,carlson2003tumor} \\ \hline
    $u$ & Missegregation rate& $10^{-2}$ & 1$\slash$cell division  & \citet{lee2016effects} \\ \hline
    $v$ & Mutation rate& $10^{-9}-10^{-7}$ &  1$\slash$cell division  & \citet{bozic2013evolutionary,loeb2001mutator} \\  \hline
    $\tilde{u}$ & Missegregation rate in the drug free environment$^\ast$& $5\times10^{-4}-2\times10^{-2}$ & 1$\slash$cell division  & \citet{shi2005chromosome,thompson2008examining} \\ \hline
    $c$ & Selection coefficient against aneuploidy in the drug free environment& $0.07$ &  1/days   & \citet{lukow2021chromosomal} \\ \hline
  \end{tabular}
\caption{\textbf{Model parameters: Melanoma.} 
Parameters from \citet{bozic2013evolutionary} consider patients with melanoma treated with the anti-cancer drug vemurfenib, in which resistance is conferred by trisomy in either Chr 2 or Chr 6.
We have modified the parameters from \citet{bozic2013evolutionary} such that sensitive and mutant division rates are $\lambda_{s} = \lambda_{m} = \log{(2)}/T\approx0.1$ instead of their value of $0.14$, where $T$ is the doubling time in the absence of cellular death obtained from \citet{rew2000cell}. For a discussion of the different interpretations of the tumor doubling times see \citet{avanzini2019cancer}. Parameters marked with $^\ast$ are not obtained from the literature.} %change 1
  \label{table1}
\end{center}
\end{table}

@@

\begin{figure}
\begin{subfigure}{0.48\textwidth}
A\\
\includegraphics[width=1\textwidth]{Figures/ProbvNPlot.pdf}
\end{subfigure}
\begin{subfigure}{0.52\textwidth}
B\\
\includegraphics[width=1\textwidth]{Figures/PDXModelProb.pdf}
\end{subfigure}
\caption{\textbf{Aneuploidy facilitates the evolutionary rescue of tumors under drug treatment.}
The probability of evolutionary rescue, $\presc$, as a function of the initial tumor size, $N$ (\cref{eq:rescue_prob}). 
Dashed vertical lines show the threshold tumor size, $N_a^*$, above which the probability is high (\cref{eq:N_a}). % change S1
\textbf{(A)} Blue dashed line: without aneuploidy ($u=0$). Black line: tolerant aneuploidy ($u=10^{-2}, \lambda_a=0.0899$). Red line: stationary aneuploidy ($u=10^{-2}, \lambda_a=0.08999$). Green line: partially resistant aneuploidy ($u=10^{-2}, \lambda_a=0.095$). Markers and error bars for averages of simulation results with $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has been rescued, and $n=100$ is the number of simulations). Parameters: $\lambda_s=0.1,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7}$.
\textbf{(B)} Comparison for melanoma and breast cancer with parameters from the literature (\Cref{table1,table2}). Black line: Melanoma A375. Blue line: Breast cancer TNBC-SA1035 clone H. Red line: Breast cancer TNBC-SA609 clone A. Green line: Breast cancer TNBC-SA535 clone H.
}
\label{rescue_prob}
\end{figure}

@@

To analyze evolutionary rescue in our model, we use the framework of \emph{multi-type branching processes} \citep{harris1963theory, weissman2009rate}. 
In the case of cancer, where all reproduction is binary fission, these branching processes are also \emph{birth-death processes} \citep{kendall1948}. % change 31
This allows us to find explicit expressions for the \emph{survival probability}: the probability that a lineage descended from a single cell does not become extinct.

@@

\caption{\textbf{Aneuploidy increases the number of rescue mutations.} Shown is the expected number of mutations, which will rescue the cancer cell population, produced through the evolutionary trajectory \textit{sensitive} $\rightarrow$ \textit{mutant} (blue line, \cref{eq:twosteplineagedirect}) or through the trajectory \textit{sensitive} $\rightarrow$ \textit{aneuploid} $\rightarrow$ \textit{mutant} (red line, \cref{eq:twosteplineage}). Dashed vertical red line represents the threshold tumor size above which evolutionary rescue is likely through aneuploidy \cref{eq:N_a} and the dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is likely to occur through direct mutation \cref{eq:N_m}. % change S1

@@

\caption{\textbf{Tumor size reduces rescue time in a non-linear fashion.}
The mean time for appearance of a resistance mutation that leads to evolutionary rescue with aneuploidy ($u>0$). 

@@

\caption{\textbf{Tumor size reduces resucrrence time in a non-linear fashion.}
The mean time for the mutant cell population to reach size $N$, where $N$ is the initial number of cancer cells.

@@

\caption{\textbf{Tumor size reduces time to detection in a non-linear fashion.}
The mean time for the mutant cell population to reach size $M$, where $M$ is the tumor detection size.
